---
title: "Conducting and Reporting ANOVAs in R"
author: "W. Joel Schneider"
date: "Psy444: Multivariate Analysis"
output: 
  slidy_presentation: 
    css: slidy.css
    fig_caption: yes
    highlight: kate
    widescreen: yes
---
<meta name="copyright" 
content="<a href='http://my.ilstu.edu/~wjschne/444/Psy444FA2015.html'>Multivariate Analysis</a>

# Load Data

```{r setup,echo=FALSE,message=FALSE}
options(digits = 2)
# Load packages
library(Hmisc)
library(ggplot2)
library(knitr)
library(dplyr)
library(readr)
library(psych)
library(broom)
library(scales)
library(tables)
library(haven)
library(tidyr)
library(lsr)
library(dummy)
library(svglite)
knit_hooks$set(familyserif = function(before, options, envir) {
    if (before) par(family = "serif")  
})
opts_chunk$set(dev = "svglite", familyserif = TRUE)

options(digits = 2)

rinline <- function(code) {
  sprintf('``` `r %s` ```', code)
}
library(pander)
panderOptions("table.split.table",Inf)
panderOptions("round",2)
panderOptions("keep.trailing.zeros",TRUE)
panderOptions("table.emphasize.rownames", FALSE)
panderOptions("table.alignment.rownames", "left")
panderOptions("missing","")
pander(head(d),caption = "First Six Rows of Data")
```


```{r MakeData, eval=FALSE, include=FALSE}
library(readr)
set.seed(4)
mu <- c(16,15,11)
sigma <- c(3.5,2,3)
n <- c(80,30,30)
lvls <- c("Control", "Low Dose", "High Dose")

d <- NULL
for (i in 1:length(mu)) {
  d <- rbind(d,data.frame(Treatment = rep(i,n[i]), Depression = rnorm(n[i],mu[i],sigma[i])))
}

d <- d[sample(1:nrow(d),nrow(d)),]

write_csv(x = d,path = "anova.csv")

b <- matrix(c(1,-1,0.5))

n <- 100
X <- matrix(rbinom(2 * n, 1, 0.5), ncol = 2)
colnames(X) <- c("A","B")
AB <- X[,1] * X[,2]
DV <- cbind(X,AB) %*% b + rnorm(n)
d2 <- data.frame(X,DV)
d2 <- d2 %>% mutate(A = factor(A, labels = c("A1", "A2")),
                   B = factor(B, labels = c("B1", "B2")))
write_csv(d2, "anova2.csv")

```

```{r ReadData}
library(readr)
d <- read_csv("http://my.ilstu.edu/~wjschne/444/anova.csv") %>% mutate(Treatment = factor(Treatment, labels = c("Control", "Low Dose", "High Dose")))
library(pander)
panderOptions("table.split.table",Inf)
panderOptions("round",2)
panderOptions("keep.trailing.zeros",TRUE)
panderOptions("table.emphasize.rownames", FALSE)
panderOptions("table.alignment.rownames", "left")
panderOptions("missing","")
pander(head(d),caption = "First Six Rows of Data")


```

# Descriptives

The `broom` package has three main functions:
* `tidy` allows us to quickly clean up output and put it into a data frame.
* `glance` puts model summary statistics into a data frame.
* `augment` generates data frames from a model and adds predicted values and standard errors.

These functions make graphing, tables, and *output-as-input* analyses very easy.

## Descriptives by Group

```{r DescriptivesByGroup}
d %>%
  group_by(Treatment) %>%
  summarize(N = length(Depression),
            Mean = mean(Depression),
            SD = sd(Depression)) %>%
  pander()


```

Using the `psych` package

```{r}
library(psych)
describeBy(d,group = "Treatment",mat = T) %>% pander()
```


Using the `tables` package

```{r}
library(tables)
tabular(Treatment ~  Depression * (mean + sd), data = d) %>%
  pander("Descriptives for Depression")
```

Make it pretty:

```{r}
tabular(Treatment ~ (`*N*` = 1) +  Heading() * Depression * 
          (Heading(`*Mean*`)* mean + (`*SD*` = sd)), data = d) %>%
  pander("Descriptives for Depression")
```



# One-Way ANOVA {.SectionSlide}

# Method 1: `aov`

This output is fine.

```{r}
m1 <- aov(Depression ~ Treatment, data = d)
pander(m1)
```

Getting it just right is harder:

```{r}
APAaov <- function(m,partial_etasq = TRUE, eta = FALSE, ...){
  require(broom) # For tidying results
  require(lsr) # For effect sizes
  require(pander) # For table output
  require(dplyr) # For aggregating data
  # Tidy model
  tm <- tidy(m)
  # Create totals
  tr <- tm %>% 
    select(df:sumsq) %>% 
    summarize(df = sum(df), sumsq = sum(sumsq)) %>% 
    mutate(statistic = NA, p.value = NA,term = "Total", meansq = sumsq/df)
  tm <- rbind(tm,tr)
  # Calculate and format effect sizes
  es <- etaSquared(m)
  es <- as.data.frame(es)  %>%  mutate(term = rownames(es))
  tm <- tm %>% left_join(es, "term")
  # Format table
  ftm <- tm %>% select(-term,-df) %>% apply(FUN = formatC, MARGIN = 2, digits = 2, format = "f")
  ftm[ftm == " NA"] <- ""
  ftm <- tm %>% select(df) %>% cbind(ftm)
  # Table headers
  rownames(ftm) <-  tm$term 
  rownames(ftm)[nrow(tm) - 1] <- "Error"
  cnames <- c("df","SS","MS","F","p")
  if (eta) cnames[length(cnames) + 1] <- "\\eta^2" 
  else ftm <- ftm %>% select(-eta.sq)
  if (partial_etasq) cnames[length(cnames) + 1] <- "\\eta_p^2" 
  else ftm <- ftm %>% select(-eta.sq.part)
  colnames(ftm) <- paste0("$",cnames,"$")
  # Output table
  pander(ftm, 
       missing = "", 
       round = 2, 
       keep.trailing.zeros = TRUE,
       ...)
}
APAaov(m1)

```

`papaja`'s `apa_print` function

```{r}
papaja::apa_print(m1)$table %>% pander()
```

`papaja::apa_print(m1)$full$Treatment`

`r papaja::apa_print(m1)$full$Treatment`

# Method 2: `lm` (ANOVA as regression)

```{r}
m2 <- lm(Depression ~ Treatment, data = d)
summary(m2)
```

```{r}
pander(anova(m2))
pander(m2)
```


## Nicer table
```{r}

sm2 <- tidy(m2) %>% select(-term)
sm2 <- sm2 %>% mutate(p.value = formatC(p.value, 2, format = "f"))
rownames(sm2) <- c("(Intercept)", "Low Dose", "High Dose")
colnames(sm2) <- paste0("$",c("b","se", "t", "p"),"$")
pander(sm2)
# Summary statistics
gm2 <- glance(m2)
pander(gm2)
gm2 <- apply(gm2, 1, formatC, digits = 2, format = "f") %>% t()
colnames(gm2) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
rownames(gm2) <- "Statistic"
pander(t(gm2))


```

# Dummy Codes

We can create them by hand:
```{r}
ddum <- d %>% mutate(dum1 = (Treatment == "Low Dose") * 1,
                  dum2 = (Treatment == "High Dose") * 1)
ddum %>% 
  head() %>% 
  pander()
```

Use the `model.matrix` function to create dummy variables. The `Intercept` column of ones is also created but it is not a dummy variable. 

```{r}
model.matrix(~Treatment,data = d) %>% 
  head() %>% 
  pander()
```

We can use the dummy package:

```{r}
library(dummy)
dums <- dummy(d)
dums %>% 
  head() %>% 
  pander()
```

# Contrasts

```{r}
lm(Depression ~ C(Treatment,contr.helmert), data = d) %>% summary()
lm(Depression ~ C(Treatment,contr.poly), data = d) %>% summary()
lm(Depression ~ C(Treatment,contr.sum), data = d) %>% summary()
lm(Depression ~ C(Treatment,contr.treatment), data = d) %>% summary()
lm(Depression ~ C(Treatment,contr.SAS), data = d) %>% summary()
```

# Homogeneity of Variance

```{r}
with(d, sjmisc::levene_test(Depression, Treatment))

plot(m2, 
     which = 1, 
     col = alpha("royalblue", 0.3),
     pch = 16, 
     cex = 0.5)

library(ggplot2)
# Set ggplot theme
themeMod <- theme_grey() + 
  theme(text = element_text(family = "serif", size = 16),
        legend.position = "none",
        axis.title.x = element_text(vjust = -0.75),
        axis.title.y = element_text(vjust = 1.75),
        plot.margin = unit(c(0.5, 0.5, 0.75, 0.75), "cm"))
theme_set(themeMod)
# Custom colors
myFill <- function(...){
  scale_fill_manual(values = alpha(c("royalblue", "firebrick", "darkorchid"),0.5),...)
}
scale_fill_discrete <- myFill

ggplot(d, aes(Depression, fill = Treatment)) + geom_density()

```

```{r, fig.height = 7, echo = -1}
par(col = alpha("royalblue", 0.3), pch = 16, bty = "n")
par(mfcol = c(2,2))
plot(m1)
```

## Bartlett Test of Homogeneity of Variances

```{r}
bartlett.test(Depression ~ Treatment, data = d)
```

## Figner-Killeen Test of Homogeneity of Variances

```{r}
fligner.test(Depression ~ Treatment, data = d) 
```

# Normality Assumption

The residuals need to be normally distributed.

## Shapiro-Wilk Normality Test

```{r}
shapiro.test(m1$residuals)
```

## Anderson-Darling Normality Test

```{r}
# A-D test on Model 1 residuals 
nortest::ad.test(m1$residuals)
```

[More on normality tests](https://rexplorations.wordpress.com/2015/08/11/normality-tests-in-r/)

# Plot Means

```{r, fig.cap="Post-Intervention Depression Disributions and Means with Error Bars"}
ggplot(d, aes(x = Treatment, y = Depression, fill = Treatment)) + 
  geom_violin(color = NA) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.1) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) 


```

# Post-hoc Comparisons

```{r}
TukeyHSD(m1)
plot(TukeyHSD(m1, conf.level = 0.95))
DescTools::ScheffeTest(m1)
```

# Two-Way ANOVA {.SectionSlide}

# Load data

```{r}
d2 <- read.csv("http://my.ilstu.edu/~wjschne/444/anova2.csv")
head(d2) %>% pander()
```

# Describe Data

## Overall stats
```{r}
tidy(d2 %>% select(DV)) %>% pander()
```

## Grouped stats

`dplyr` functions can create our data easily but the formatting won't be exactly right.

```{r}
d2 %>% 
  group_by(A,B) %>% 
  summarise(N = length(DV),
            Mean = mean(DV),
            SD = sd(DV)) %>% 
  pander()
```

The `tables` package is extremely flexible. The formula means:

`Rows ~ Columns`

`A * B` means create all combinations of `A` and `B`.

`mean + sd` means create two columns for means and standard deviations.

```{r}
library(tables)
tabular(A * B ~  DV * (mean + sd), 
        data = d2) %>% 
  pander()
```

```{r}
tabular(DV * (mean + sd) ~ (A * B), 
        data = d2) %>% 
  pander()

tabular(Heading() * A ~  
          Heading() * B * Heading() * DV * (mean + sd), 
        data = d2) %>% 
  pander()
```


Let's format the table to make it just right.

```{r}

fmt <- function(x) htmlNumeric(formatC(x, 
                                       digits = 2, 
                                       format = "f"), 
                               leftpad = FALSE)
stderr <- function(x) sd(x)/sqrt(length(x))

fmtLatex <- function(x) paste0("$", x, "$")


tabular((Factor(A, levelnames = c("A~1~","A~2~")) + 1) *
        (Factor(B, levelnames = c("B~1~","B~2~")) + 1) ~ 
        (`*n*` = 1) + 
        Heading() * DV * ((`*M*` = mean) + 
                          (`*SD*` = sd) + 
                          (`*SE*` = stderr)) * 
        Format(fmt()), 
        data = d2) %>% 
  pander()
```

# Factorial ANOVA

```{r}
mFact <- lm(DV ~ A * B, data = d2)
pander(anova(mFact))
pander(mFact)

```

# Plot Means

```{r}
pTW <- ggplot(d2, aes(x = A, y = DV, fill = A)) 

pTW + 
  geom_violin(color = NA) + 
  stat_summary(fun.data = "mean_cl_normal") + 
  facet_grid(. ~ B)

TukeyHSD(aov(mFact))
par(mar = c(5,7,3,1))
plot(TukeyHSD(aov(mFact)),las = 1)
```

# Repeated-Measures ANOVA {.SectionSlide}

# Load Data

```{r}
d3 <- haven::read_spss("http://my.ilstu.edu/~wjschne/138/TherapyOutcome.sav")
d3 %>% head() %>% pander()
# Rename variables to make them easier to restructure
# Also mutate Therapy and PersonID into factors so that ggplot knows what to do with them.
d3 <- d3 %>% 
  rename(Therapy = TherapyType,
         Aggression_1 = aggression1,
         Aggression_2 = aggression2,
         Empathy_1 = empathy1,
         Empathy_2 = empathy2) %>% 
  mutate(Therapy = as_factor(Therapy),
         PersonID = factor(PersonID))

d3 %>% head() %>% pander()
```

# Scatter Plots

```{r}
plot(d3 %>% select(Aggression_1:Empathy_2))
```

# Restructure Data into a Tidy Data Frame

First, we will do this in steps. However, at the end, we will re-do it all in one step.

```{r, warning=FALSE}
library(tidyr)
# Put variable names into a column called VariablesTime
d3r <- d3 %>% gather("VariablesTime", "Values", Aggression_1:Empathy_2)
d3r

# Separate VariablesTime into Variables and Time
d3r <- d3r %>% separate(VariablesTime, c("Variables","Time")) 
d3r

# Make Time a factor
d3r <- d3r %>% mutate(Time = as_factor(Time))
d3r

# Spread Variables into Aggression and Empathy columns
d3r <- d3r %>% spread(Variables, Values) 
d3r

# Do all these changes in one step
d3r <- d3 %>% 
  gather("VariablesTime", "Values", Aggression_1:Empathy_2) %>%
  separate(VariablesTime, c("Variables","Time")) %>% 
  mutate(Time = as_factor(Time)) %>% 
  spread(Variables, Values)
d3r
```

# Descriptives

```{r, results='asis'}
library(tables)
tabular((Therapy + 1) * (Time + 1) ~ 
          (N = 1) + 
          (Aggression + Empathy) * (mean + sd), 
        data = d3r) %>% pander()
```

# Plot Distributions

## Aggression

```{r}
ggplot(d3r, aes(Time, Aggression)) +
  geom_violin() + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) + 
  stat_summary(fun.y = "mean", geom = "point") + 
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.1) + 
  facet_grid(. ~ Therapy, labeller = label_both)
```

## Empathy

```{r}
ggplot(d3r, aes(Time, Empathy)) +
  geom_violin() + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) + 
  stat_summary(fun.y = "mean", geom = "point") + 
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.1) + 
  facet_grid(. ~ Therapy, labeller = label_both)
```

# Analyses

Make sure that `PersonID` is a factor.

```{r}
# These options make sure that the results are the same as that of SPSS
options(contrasts = c("contr.sum","contr.poly"))


m3a <- aov(Aggression ~ Therapy * Time + Error(PersonID/Time), data = d3r)
m3a %>% tidy() %>% pander()
d3 <- d3 %>% mutate(dAgg = Aggression_2 - Aggression_1)
aov(dAgg ~ Therapy, d3) %>% summary
m3e <- aov(Empathy ~ Therapy * Time + Error(PersonID/Time), data = d3r)
```

# Repeated-Measures ANOVA as Regression (with Random Intercepts)

```{r}
# lme4 does multilevel modeling
library(lme4)
# lmerTest makes lme4 give p-values
library(lmerTest)
# (1 | PersonID) means that each unique PersonID gets its own intercept
m3a2 <- lmer(Aggression ~ Time * Therapy + (1 | PersonID), data = d3r) 
summary(m3a2)
anova(m3a2)

plot(m3a2)
qqmath(m3a2, id = 0.05)
# Make a data frame with fitted values from model
am2a2 <- augment(m3a2)

ggplot(am2a2, aes(Time, Aggression, color = PersonID, group = PersonID)) + geom_line()

ggplot(am2a2, aes(Time, Aggression)) +
  geom_line(aes(group = PersonID, color = PersonID)) +
  facet_grid(~Therapy)

ggplot(am2a2, aes(Time, .fitted)) + geom_line(aes(group = PersonID, color = PersonID), alpha = 0.5) + facet_grid(~Therapy)


am2a2 <- am2a2 %>% mutate(Timej = as.numeric(Time) + rnorm(nrow(am2a2), mean = 0, sd = 0.02))
glimpse(am2a2)
ggplot(am2a2, aes(Time, Aggression))  + 
  geom_violin(color = NA) + 
  geom_segment(aes(x = Timej, 
                   xend = Timej, 
                   y = .fitted, 
                   yend = Aggression, 
                   group = PersonID, 
                   color = PersonID), 
               arrow = arrow(type = "closed", 
                             length = unit(0.15,"cm")), 
               alpha = 0.5) + 
  geom_line(aes(x = Timej, 
                y = .fitted, 
                group = PersonID, 
                color = PersonID), 
            alpha = 0.5) + 
  geom_line(aes(y = .fixed, group = 1),lwd =1) +
  facet_grid(~Therapy)
```



