---
title: "Canonical Correlation Analysis in R"
author: "W. Joel Schneider"
date: "Psy444: Multivariate Analysis"
params: 
  fast: TRUE
output: 
  slidy_presentation: 
    css: slidy.css
    fig_caption: yes
    highlight: kate
    widescreen: yes
    footer: <a href = 'http://my.ilstu.edu/~wjschne/444/Psy444FA2015.html'>Multivariate Analysis</a>
bibliography: Loaded.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ev <- T
library(magrittr)
library(candisc)
library(tables)
library(ggplot2)
library(knitr)
library(broom)
library(pander)
library(ztable)
library(dplyr)

knit_hooks$set(familyserif = function(before, options, envir) {
    if (before) par(family = "serif")  
})
opts_chunk$set(dev = "svglite", familyserif = TRUE)

panderOptions("table.split.table",Inf)
panderOptions("round",2)
panderOptions("keep.trailing.zeros",TRUE)
panderOptions("table.emphasize.rownames", FALSE)
panderOptions("table.alignment.rownames", "left")
panderOptions("table.style","rmarkdown")
panderOptions("missing","")
pvalueAPA <- function(p, inline = FALSE, mindigits = 2, maxdigits = 3){
   fp <- function(x){
     if (is.na(x)) return(NA)
      p.round <- ifelse(x > 0.5 * 10 ^ (-1 * mindigits),mindigits,maxdigits)
  if (x > 0.5 * 10 ^ (-1 * p.round)) {
    paste0(ifelse(inline,"$p=", ""),
           sub(pattern = "0.", 
               replacement = ".", 
               formatC(x, p.round, format = "f")),
           ifelse(inline,"$", ""))
    } else {
      paste0(ifelse(inline, "$p<","<"),
             sub(pattern = "0.", 
                 replacement =  ".",
                 10 ^ (-1 * maxdigits)),
             ifelse(inline,"$",""))
    }
  }
  sapply(p, fp)
}
manovaTable <- function(m, caption = "MANOVA Table", test = "Pillai"){
  require(broom)
  require(pander)
  require(dplyr)
  testName <- switch(test,
                     Pillai = "Pillai's Trace", 
                     Wilks = "Wilks' $\\Lambda$",
                     `Hotelling-Lawley` = "Hotelling-Lawley's Trace",
                     Roy = "Roy's Largest Root")
  m %>% manova %>% 
    tidy(test = test) %>% 
    mutate(p.value = pvalueAPA(p.value),
           term = gsub(x = term, pattern = ":"," $\\\\times$ ") ) %>% 
    set_colnames(c("Factor",
                   "df",
                   testName,
                   "$F$",
                   "$df_1$",
                   "$df_2$",
                   "$p$")) %>% 
    pander(caption = caption, 
           round = 2, 
           split.tables = Inf, 
           style = "rmarkdown", 
           emphasize.rownames = FALSE, 
           keep.trailing.zeros = TRUE, 
           table.alignment.rownames = "left", 
           missing = "")
}
manovaUnivariateTable <- function(m, caption = "Univariate Analyses"){
  require(broom)
  require(pander)
  require(dplyr)
  require(car)
  m %>% tidy %>% 
    mutate(response = ifelse(response == Lag(response),
                             NA,
                             response),
           p.value = pvalueAPA(p.value),
           term = gsub(x = term, pattern = ":"," $\\\\times$ "))  %>% 
    mutate(term = gsub(x = term, pattern = "[[:punct:]]","")) %>% 
    set_colnames(c("Outcome",
                   "Predictor",
                   "$b$",
                   "$\\sigma_b$",
                   "$t$",
                   "$p$")) %>% 
    pander(caption = "Univariate Analyses")
}

CanCorTable <- function(x) {
  require(candisc)
  require(pander)
  require(magrittr)
  if (class(x) != "cancor") return("Object must be of class cancor from the candisc package.")
  k <- length(x$cancor)
  Eigenvalue = x$cancor ^ 2 / (1 - x$cancor ^ 2)
  EigPercent <- 100 * Eigenvalue / sum(Eigenvalue)
  data.frame(Variate = 1:k, 
           CanR = x$cancor, 
           CanRSq = x$cancor ^ 2, 
           Eigenvalue, 
           Percent = EigPercent, 
           CumulativePercent = cumsum(EigPercent)) %>% 
  set_colnames(c("Variate", 
                 "Canonical $R$",
                 "Canonical $R^2$",
                 "Eigenvalues",
                 "Percent",
                 "Cumulative Percent")) %>% 
  pander(caption = "Canonical Correlation Analysis")
  }

CanCorTestTable <- function(x) {
  require(candisc)
  require(pander)
  require(magrittr)
  if (class(x) != "cancor") return("Object must be of class cancor from the candisc package.")
  k <- length(x$cancor)
  cbind(Variate = 1:k, Wilks(x)) %>% 
    mutate(p.value = pvalueAPA(p.value)) %>%
    set_colnames(c("Variate", 
                   "Canonical $R$", 
                   "Wilks $\\Lambda$", 
                   "$F$", 
                   "$df_1$",
                   "$df_2$",
                   "$p$")) %>% 
    pander(caption = "Test of H~0~: The canonical correlations in the current row and all that follow are zero.")
  }

CanCorCoefTable <- function(x, 
                            type = c("raw",
                                     "standardized",
                                     "structure"), 
                            xNames = x$names$X, 
                            yNames = x$names$Y, 
                            xSet = "X", 
                            ySet = "Y",
                            boldthreshold = 0.3,
                            digits = 2) {
  require(candisc)
  require(pander)
  require(magrittr)
  bolder <- function(p, 
                     boldthreshold = boldthreshold, 
                     digits = digits) {
    p0 <- formatC(p,digits = digits, format = "f")
    if (abs(p) > boldthreshold) paste0("**",p0 ,"**") else p0}
  
  k <- length(x$cancor)
  XTitleRow <- matrix(rep(NA,k),nrow = 1) %>%
    set_rownames(paste0("**",xSet,"**"))
  YTitleRow <- matrix(rep(NA,k),nrow = 1) %>% set_rownames(paste0("**",ySet,"**"))
if (type == "structure") {
  xcoef <- x$structure$X.xscores %>%
    set_rownames(xNames) %>% 
    apply(MARGIN = c(1,2),
          bolder, 
          boldthreshold = boldthreshold, 
          digits = digits)
  ycoef <- x$structure$Y.yscores %>%
    set_rownames(yNames) %>% 
    apply(MARGIN = c(1,2),
          bolder, 
          boldthreshold = boldthreshold, 
          digits = digits)
  cap <- "Canonical Structure Coefficients"
} 
 
if (type == "raw") {
  xcoef <- x$coef$X %>%
    set_rownames(xNames)
  ycoef <- x$coef$Y %>%
    set_rownames(yNames)
  cap = "Canonical Function Coefficients"
}
  
if (type == "standardized") {
  xcoef <- coef(x, 
                type = "x", 
                standardize = TRUE) %>% 
    set_rownames(xNames) %>% 
    apply(MARGIN = c(1,2), 
          bolder, 
          boldthreshold = boldthreshold, 
          digits = digits)
  ycoef <- coef(x, 
                type = "y", 
                standardize = TRUE) %>% 
    set_rownames(yNames) %>% 
    apply(MARGIN = c(1,2), 
          bolder, 
          boldthreshold = boldthreshold, 
          digits = digits)
  cap <- "Standardized Canonical Function Coefficients"
  }
rbind(XTitleRow,
      xcoef,
      YTitleRow,
      ycoef) %>%
  set_colnames(paste0("CV",1:k)) %>% 
  pander(caption = cap,
         justify = paste0("l",
                          paste0(rep("r",k),
                                 collapse = "")))
  }
```

```{r MakeData, include = FALSE}
library(lavaan)

O <- paste0("O =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * O",1:5),collapse = "+"))
C <- paste0("C =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * C",1:5),collapse = "+"))
E <- paste0("E =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * E",1:5),collapse = "+"))
A <- paste0("A =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * A",1:5),collapse = "+"))
N <- paste0("N =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * N",1:5),collapse = "+"))

Real <- paste0("Real =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Real",1:5),collapse = " + "))
Invest <- paste0("Invest =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Invest",1:5),collapse = " + "))
Art <- paste0("Art =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Art",1:5),collapse = " + "))
Soc <- paste0("Soc =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Soc",1:5),collapse = " + "))
Ent <- paste0("Ent =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Ent",1:5),collapse = " + "))
Conv <- paste0("Conv =~ ", paste0(paste0(round(runif(5,0.6,0.9),2)," * Conv",1:5),collapse = " + "))

vNames <- expand.grid(c("O","C","E","A","N","Real","Invest","Art","Soc","Ent","Conv"),1:5) %>% apply(1,paste, collapse = "") 
thresholds <- paste0(vNames,"| -1.5*t1 + -0.5*t2 + 0.5*t3 + 1.5*t4",collapse = "\n")
m <- paste("Real ~ -0.3 * O + -0.25 * E",
           "Invest ~ 0.6 * O",
           "Art ~ 0.3 * O",
           "Soc ~ 0.5 * E + 0.5 * A + -0.3 * O",
           "Ent ~ 0.6 * E + -0.3 * A",
           "Conv ~ -0.6 * O + -0.4 * E",
  O,C,E,A,N,
  Real,Invest,Art,Soc,Ent,Conv, 
  thresholds, 
  sep = "\n")
d <- simulateData(model = m, sample.nobs = 500, standardized = TRUE)

```



# Multiple Regression

```{tikz MR, engine.opts = list(convert = 'C:\\Program Files\\ImageMagick-6.9.2-Q16\\convert.exe', convert.opts = '-density 300'), echo = FALSE, eval = ev, dev = "png"}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{decorations.text}
\usetikzlibrary{arrows,shapes,backgrounds, shadows,fadings,trees}
\usetikzlibrary{calc}
\begin{tikzpicture}[scale=1,  latent/.style={ellipse, ultra thick ,draw=black,minimum width=1.5cm,minimum height=1.5cm,inner sep=0mm,font=\Large},
error/.style={ellipse,thick, draw=black,inner sep=0mm, minimum width=0.5cm,minimum height=0.5cm,font=\Large},
ob/.style={rectangle,draw=black, thick,inner sep=2mm, minimum width=1cm, minimum height=0.75cm, align=center, rounded corners=0.1cm,font=\Large},
post/.style={->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large},
cov/.style={<->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large, bend left=50},
variance/.style={<->,  >=latex', thick, bend left=245, looseness=6,shorten >=2pt,shorten <=2pt, font=\Large},
upvariance/.style={<->,  >=latex', thick, shorten >=2pt,shorten <=2pt, font=\Large,bend left = 115,looseness=4},
label/.style={fill=white,circle,inner sep = 0mm,pos=0.475}]

\node[latent, minimum height=1cm] (cv1) at (0,0) {$\hat{Y}$};
\node[latent, minimum height=1cm] (cv2) [right= of cv1] {$Y$};
\node[ob] (cv12) [left = of cv1] {$x_2$};
\node[ob] (cv11) [above = of cv12] {$x_1$};
\node[ob] (cv13) [below = of cv12] {$x_3$};
\node[ob] (cv22) [right = of cv2] {$Y$};
\path[cov] (cv1) to (cv2); 

\foreach \m in {1,2,3}{
\path[post] (cv1\m) to (cv1);
}

\path[post] (cv22) to (cv2);
\end{tikzpicture}
```

# Canonical Correlation Analysis

```{tikz CCA, engine.opts = list(convert = 'C:\\Program Files\\ImageMagick-6.9.2-Q16\\convert.exe', convert.opts = '-density 300'), echo = FALSE, eval = ev, dev = 'png'}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{decorations.text}
\usetikzlibrary{arrows,shapes,backgrounds, shadows,fadings,trees}
\usetikzlibrary{calc}
\begin{tikzpicture}[scale=1,  latent/.style={ellipse, ultra thick ,draw=black,minimum width=1.5cm,minimum height=1.5cm,inner sep=0mm,font=\Large},
error/.style={ellipse,thick, draw=black,inner sep=0mm, minimum width=0.5cm,minimum height=0.5cm,font=\Large},
ob/.style={rectangle,draw=black, thick,inner sep=2mm, minimum width=1cm, minimum height=0.75cm, align=center, rounded corners=0.1cm,font=\Large},
post/.style={->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large},
cov/.style={<->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large, bend left=50},
variance/.style={<->,  >=latex', thick, bend left=245, looseness=6,shorten >=2pt,shorten <=2pt, font=\Large},
upvariance/.style={<->,  >=latex', thick, shorten >=2pt,shorten <=2pt, font=\Large,bend left = 115,looseness=4},
label/.style={fill=white,circle,inner sep = 0mm,pos=0.475}]

\node[latent, minimum height=1cm] (cv1) at (0,0) {$\xi$};
\node[latent, minimum height=1cm] (cv2) [right= of cv1] {$\eta$};
\node[ob] (cv12) [left = of cv1] {$x_2$};
\node[ob] (cv11) [above = of cv12] {$x_1$};
\node[ob] (cv13) [below = of cv12] {$x_3$};
\node[ob] (cv22) [right = of cv2] {$y_2$};
\node[ob] (cv21) [above = of cv22] {$y_1$};
\node[ob] (cv23) [below = of cv22] {$y_3$};
\path[cov] (cv1) to (cv2); 

\foreach \m in {1,2,3}{
\path[post] (cv1\m) to (cv1);
\path[post] (cv2\m) to (cv2);
}
\end{tikzpicture}
```

+ $r_{\xi\eta}$ is the *canonical correlation*.
+ The values associated with the straight arrows are *canonical function coefficients*. These can be unstandardized or standardized.
+ The correlation between each $x$ and $\xi$ (or each $y$ and $\eta$) is a *canonical structure coefficient* (also known as a *canonical loading*).
+ The correlation of each variable with its opposite canonical variate is the *canonical cross-loading*.

# Canonical Correlation Analysis

```{tikz CCA2, engine.opts = list(convert = 'C:\\Program Files\\ImageMagick-6.9.2-Q16\\convert.exe', convert.opts = '-density 300'), echo = FALSE, eval = ev, dev = "png"}

\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{decorations.text}
\usetikzlibrary{arrows,shapes,backgrounds, shadows,fadings,trees}
\usetikzlibrary{calc}
\begin{tikzpicture}[scale=1,  latent/.style={ellipse, ultra thick ,draw=black,minimum width=1cm,minimum height=1cm,inner sep=0mm,font=\Large},
error/.style={ellipse,thick, draw=black,inner sep=0mm, minimum width=0.5cm,minimum height=0.5cm,font=\Large},
ob/.style={rectangle,draw=black, thick,inner sep=2mm, minimum width=1cm, minimum height=1cm, align=center, rounded corners=0.1cm,font=\Large},
post/.style={->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large},
cov/.style={<->,draw,shorten >=2pt,shorten <=2pt,>=latex', very thick, font=\Large, bend left=50},
variance/.style={<->,  >=latex', thick, bend left=245, looseness=6,shorten >=2pt,shorten <=2pt, font=\Large},
upvariance/.style={<->,  >=latex', thick, shorten >=2pt,shorten <=2pt, font=\Large,bend left = 115,looseness=4},
label/.style={fill=white,circle,inner sep = 0mm,pos=0.475}]
\node[ob] (x1) at (0,0) {$x_1$};



\foreach \i/\j in {1/2,2/3,3/4}{
  \node[ob] (x\j) [below = of x\i] {$x_\j$};
  \node[ob] (y\i) [right = 8cm of $(x\i)!0.5!(x\j)$] {$y_\i$};
}

\foreach \i in {1,2,3}{
  \node[latent] (eta\i) [left = 2cm of y\i] {$\eta_\i$};
  \node[latent] (xi\i) [left = of eta\i] {$\xi_\i$};
  \path[cov] (xi\i) to (eta\i);
  \foreach \k in {1,2,3}{
     \path[post] (x\k) to (xi\i);
     \path[post] (y\k) to (eta\i);
  }
}

\foreach \i in {1,2,3} \path[post] (x4) to (xi\i);
\end{tikzpicture}
```

- $\xi_1$ and $\eta_1$ are the first *canonical variate pair*.
- $\xi_2$ and $\eta_2$ are the second *canonical variate pair*.
- $\xi_3$ and $\eta_3$ are the second *canonical variate pair*.

## How many pairs?

#. Count the number of variables on the left
#. Count the number of variables on the right.
#. Choose the smaller number.

# Data

How are personality and career interests related?

## Personality

#. Openness to Experience (0)
#. Conscientiousness (C)
#. Extraversion (E)
#. Agreeablness (A)
#. Neuroticism (N)

## Career Interests

#. Realistic
#. Investigative
#. Artistic
#. Social
#. Enterprising
#. Conventional

```{tikz RIASEC, engine.opts = list(convert = 'C:\\Program Files\\ImageMagick-6.9.2-Q16\\convert.exe', convert.opts = '-density 300'), echo = FALSE, eval = ev, dev = "png"}
\begin{tikzpicture};

\foreach \i in {0,60,120,180,240,300} {
\node[inner sep=0] (x\i) at (\i:5cm) {};
\fill (x\i) circle (2pt);}

\foreach \i/\j in {0/60,60/120,120/180,180/240,240/300,300/0} {
\draw[-] (x\i) to (x\j);}
\node[rotate =30] at (120:5.3cm){Realistic};
\node[rotate = -30] at (60:5.3cm){Investigative};
\node[rotate = -90] at (0:5.3cm){Artistic};
\node[rotate = 30] at (300:5.3cm){Social};
\node[rotate = -30] at (240:5.3cm){Enterprising};
\node[rotate = -90] at (180:5.3cm){Conventional};
\end{tikzpicture};


```

```{r}
glimpse(d)
```


# Make Composite Scores

```{r}
d <- d %>% rowwise %>% mutate(O = mean(O1:O5),
                              C = mean(C1:C5),
                              E = mean(E1:E5),
                              A = mean(A1:A5),
                              N = mean(N1:N5),
                              Real = mean(Real1:Real5),
                              Invest = mean(Invest1:Invest5),
                              Art = mean(Art1:Art5),
                              Soc = mean(Soc1:Soc5),
                              Ent = mean(Ent1:Ent5),
                              Conv = mean(Conv1:Conv5))
glimpse(d)
```

```{r, include=FALSE}
write.csv(d,"OCEANRIASEC.csv")
```


# Means

```{r}
tabular((Heading("Openness") * O + Heading("Conscientiousness") * C + Heading("Extraversion") * E + Heading("Agreeableness") * A + Heading("Neuroticism") * N + Heading("Realistic") * Real + Heading("Investigative") * Invest + Heading("Artistic") * Art + Heading("Social") * Soc + Heading("Enterprising") * Ent + Heading("Conscientiousness") * Conv) ~ (Heading("Mean") * mean + Heading("SD") * sd), data = d) %>% pander
```

# Correlations
```{r, results='asis', echo=FALSE}
library(ztable)
z <- d %>% 
  select(O:Conv) %>% 
  cor %>% 
  ztable(type = "html", caption = "Correlations")  
for (i in 2:6) z <- addCellColor(z,
                                 rows = i,
                                 cols = 2:6,
                                 color = "mediumslateblue")
for (i in 7:12) z <- addCellColor(z,
                                  rows = i,
                                  cols = 2:6,
                                  color = "mediumorchid")
for (i in 2:6) z <- addCellColor(z,
                                 rows = i,
                                 cols = 7:12,
                                 color = "mediumorchid")
for (i in 7:12) z <- addCellColor(z,
                                  rows = i,
                                  cols = 7:12,
                                  color = "utahcrimson")
z
```

$$\boldsymbol{R = \color{crimson}{R_{yy}^{-1}}\color{mediumorchid}{R_{yx}}\color{mediumslateblue}{R_{xx}^{-1}}\color{mediumorchid}{R_{xy}}}$$

# CCA in R

```{r}
RR <- cor(d %>% select(O:Conv))
Rxx <- RR[1:5,1:5]
Rxy <- RR[1:5,6:11]
Ryx <- RR[6:11,1:5]
Ryy <- RR[6:11,6:11]

R <- solve(Rxx) %*% Rxy %*% solve(Ryy) %*% Ryx
eigR <- eigen(R)
eigR$values %>% sqrt

stats::cancor(d %>% select(O:N), d %>% select(Real:Conv))
```

# CCA using the `candisc` package

```{r}
library(candisc)
cca <- candisc::cancor(d %>% select(O:N), d %>% select(Real:Conv))
cca %>% summary

CanCorTable(cca)
CanCorTestTable(cca)

cca$coef$X %>% pander
coef(cca, type = "x", standardize = TRUE) %>% pander
cca$coef$Y %>% pander
coef(cca, type = "y", standardize = TRUE) %>% pander
cca$scores$X %>% head
cca$scores$Y %>% head

plot(cca, which = 1)
plot(cca, which = 2)
plot(cca, which = 3)
plot(cca, which = 4)
plot(cca, which = 5)

cor(cbind(cca$scores$X,cca$scores$Y)) %>% pander
cca$structure$X.xscores %>% pander
cca$structure$X.yscores %>% pander
cca$structure$Y.yscores %>% pander
cca$structure$Y.xscores %>% pander


```

# Interpreting CCA

```{r}
PersonalityNames <- c("Openness",
                      "Conscientiousness",
                      "Extraversion",
                      "Agreeableness",
                      "Neuroticism")
InterestNames <- c("Realistic",
                   "Investigative",
                   "Artistic",
                   "Social",
                   "Enterprising",
                   "Conventional")
CanCorCoefTable(cca,
                type = "standardized",
                xNames = PersonalityNames,
                yNames = InterestNames,
                xSet = "Personality",
                ySet = "Interests")
CanCorCoefTable(cca,
                type = "structure",
                xNames = PersonalityNames,
                yNames = InterestNames,
                xSet = "Personality",
                ySet = "Interests")

heplot(cca)
heplot(cca, which = c(1,3))
heplot(cca, which = c(2,3))      
```


# Redundancy Analysis

Proportion of variance in X explained by Y and vice-versa

```{r}
redundancy(cca)
```

# Multivariate Multiple Regression

```{r}
mxy <- lm(cbind(Real, Invest, Art, Soc, Ent, Conv) ~ O + C + E + A + N, d)
mxy %>% manovaTable
mxy %>% manovaUnivariateTable
library(lm.beta)
lm.beta(mxy) %$% standardized.coefficients %>% pander
pairs(mxy)
```

# CCA in SPSS

`MANOVA` (available in syntax only).


