---
title: "Conducting and Reporting *t*-tests in R"
author: "W. Joel Schneider"
date: "Psychology 444<br>Multivariate Analysis"
output: 
  slidy_presentation: 
    css: slidy.css
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: kate
    widescreen: yes
---


# Data Management{.SectionSlide}

```{r setup, include=FALSE}
# Install these packages if they are missing
list.of.packages <- c("ggplot2", "knitr", "dplyr", "readr", "ggthemes", "ztable", "stargazer", "sjPlot", "sjmisc","Hmisc","tidyr","shape","scales")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if (length(new.packages)) install.packages(new.packages)

# Load packages
library(knitr)
library(Hmisc)
library(dplyr)
library(ggplot2)
library(readr)
library(ztable)
library(stargazer)
library(sjPlot)
library(sjmisc)
library(tidyr)
library(shape)
library(scales)
library(svglite)
opts_chunk$set(
  dev = "svglite"
)
options(digits = 2)

rinline <- function(code) {
  sprintf('``` `r %s` ```', code)
}
```


```{r MakeData, include=FALSE}
library(readr)
set.seed(1)
mu1 <- 10
sigma1 <- 3
mu2 <- 13
sigma2 <- 3
n1 <- 58
n2 <- 87
x1 <- rnorm(n = n1,mean = mu1,sd = sigma1)
x2 <- rnorm(n = n2,mean = mu2,sd = sigma2)
x <- c(x1,x2)
group <- factor(c(rep(1,n1),rep(2,n2)),labels = c("Group 1","Group 2"))
d <- data.frame(IV = group,DV = x)
write_csv(x = d,path = "ttest.csv")

d2 <- data.frame(id = 1:n1, x_1 = x1, x_2 = x2[1:n1])
write_csv(x = d2,path = "pairedttest.csv")
```


# Import Data

Most data can be stored in the .csv format (Comma-separated values).

The base R `read.csv` function works great. The `readr` package has a faster function called `read_csv` with better defaults.

```{r import}
d <- read.csv("http://my.ilstu.edu/~wjschne/444/ttest.csv")
```

* `IV`: Independent Variable
    + Group 1
    + Group 2
* `DV`: Dependent Variable

# Other Packages for Importing Data

* The `haven` package will read and write data created in SPSS, SAS, and Stata. The `Hmisc` and `foreign` packages have similar functions.
* The `readxl` package reads data stored in Excel, as does the `xlsx` package.
* There are many packages that read from databases, including `RODBC`, `DBI`, and `sqldf`.

[Video tutorial on importing data into R](https://vimeo.com/130548869)

# View Data

To see the first few cases use the `head` function. The `tail` function displays the last few cases.

```{r headtail}
head(d)
tail(d)
```

To see the entire data set, use the `View` function.

```{r viewdata, eval = FALSE}
View(d)
```

The `dplyr` package has a nifty `glimpse` function for viewing a little bit of the each variable.

```{r glimpsedata}
library(dplyr)
glimpse(d)
```

Here we see the variable names, the data types,and the first few data points.

## Data types in R

* lgl: Logical data  (i.e., `TRUE` and `FALSE`).
* int: Integer data (e.g. `1L`, `2L`, ...0)
* dbl: Numeric/Double-precision floating-point (e.g. `2.38`)
* cmpl: Complex (e.g., `6 + 3i`)
* fctr: Factor = nominal data stored as numbers (e.g., `1 = "Male"`)



# Independent-Samples *t*-tests {.SectionSlide}

$$H_0: \mu_0=\mu_1$$

$$H_1: \mu_0\ne\mu_1$$



# Running the Independent Samples *t*-test

Unlike most programs, the `t.test` function assumes *unequal* variances by default. I think this is a good idea.

```{r varunequal}
with(d, t.test(DV ~ IV))
```

Set `var.equal` to `TRUE` for equal variances.

```{r varequal}
with(d, t.test(DV ~ IV, var.equal = TRUE))
```

# Levene's Test for Homogeneity of Variance

```{r levene}
library(sjmisc)
with(d, levene_test(DV, IV))
```

The `leveneTest` function from the `car` package is more widely known but I like the `sjmisc` output better. 

Other homogeneity of variance tests: `fligner.test` and `bartlett.test`

```{r homogeneity}
fligner.test(DV ~ IV, data = d)
bartlett.test(DV ~ IV, data = d)
```


# Descriptives

We will make extensive use of the `dplyr` package for data manipulation. It also allows us to use the pipe operator: `%>%`. It takes the output of the last function and inserts it into the first slot of the next function. Think of it as the "Then-Do-This" function. It makes code much easier to write and read. The RStudio shortcut for `%>%` is Ctrl+Shift+M (&#8984;+Shift+M for Mac).

Let's calculate summary statistics for each group. 

```{r groupsummary}
library(dplyr)
# Summarize data by groups
md <- d %>% # "Start with the data set we imported, d 
  group_by(IV) %>% # Then group d by IV
  summarize(N = length(DV), # Then summarize each group
            Mean = mean(DV),
            SD = sd(DV),
            SE = SD/sqrt(N)) 

# Display md
md 
```

Not pretty, is it? It isn't supposed to be. Statistical output in R is always raw text. This seeming weakness is actually one of the great advantages of using R. Raw text can be formatted (by R or by other programs) to create publication-ready output without having to be cut and pasted by you. 

RStudio, under the hood, is taking your statistical output, combining it with your RMarkdown document, and creating a Markdown document using the `knitr` package. Then it hands the Markdown document to Pandoc, which converts it to one of many possible formats, including MS Word, HTML, or .pdf. 

This slideshow was created in RMarkdown and converted by Pandoc to an HTML5 slide format called Slidy.

More complex formatting is available using what RStudio calls "R Sweave" documents (.Rnw). The `knitr` package "knits" R code and $\LaTeX$ to create a .tex file, which is converted to a .pdf document using pdfLaTeX. Actually, pdfLatex handles the last step of the conversion from RMarkdown to .pdf format.

Other R packages, including `pander` and `ReportR`, can create complex documents in MS Word or html.

# Display data in tables

For simple tables, the `kable` function in the `knitr` package works well.

```{r kabletable}
library(knitr)
kable(md, digits = 2)
```

If we are going to use the `kable` function just once and loading the whole `knitr` library seems like overkill, you can use the `::` function. 

```{r justkable, eval= FALSE}
knitr::kable(md, digits = 2)
```

The `::` function allow us to use any function from any installed package, whether it is loaded or not. Sometimes functions from different packages have the same name. Using the `::` function removes the ambiguity.

We would also like statistics for the entire data set. Note that I added `IV = "Total"` so that I can later combine the total statistics with the group statistics in one table.

```{r summarizedplry}
# Summarize whole data set
td <- d %>% summarize(IV = "Total",
                      N = length(DV), 
                      Mean = mean(DV),
                      SD = sd(DV),
                      SE = SD/sqrt(N))
# Display td
td 
# Combine data into one object
dd <- rbind(md,td)
# Display combined table
kable(dd, digits = 2)
```

For more complex tables, we can use many other packages, including `xtable`, `ztable`, `pander`, `htmlTable`, `stargazer`, `sjPlot`, `memisc`, and `tables`.

# Bar Plot

```{r barplotggplot}
# Load ggplot2 library
library(ggplot2)
# Set ggplot defaults
themeMod <- theme_grey() + 
  theme(text = element_text(family = "serif", size = 10),
        legend.position = "none", axis.text.y = element_text(hjust = 1))
theme_set(themeMod)
# Custom colors 
myFill <- function(...){
  scale_fill_manual(values = alpha(c("royalblue","firebrick"),0.5),...)
}

scale_fill_discrete <- myFill

# Tell gpplot what data to load
# what to put on x and y axis
p <- ggplot(data = d, aes(y = DV, 
                          x = IV,
                          fill = IV)) 
p + stat_summary(fun.y = "mean", 
                 geom = "bar") 
```

# Error Bars

```{r ggerrorbars}
p + stat_summary(fun.y = "mean", 
                 geom = "bar") + 
  stat_summary(fun.data = "mean_cl_normal", 
               geom = "errorbar", 
               width = 0.1)

```

# Box Plot

```{r plot_t}
p + geom_boxplot() 
```

The line in the box is the median, not the mean. 

The box is the interquartile range. 

The whiskers are roughly the middle 95% CI.

The points are outliers.

# Density Plot

```{r density}
ggplot(d, aes(x = DV, fill = IV)) + 
  geom_density(color = NA) 
```

# Violin Plot

```{r violin}
p + geom_violin(color = NA) 
```

The violin plot is a centered vertical density plot.

# Combined Violin and Notched Boxplot

```{r violinnotch}
p + geom_violin(alpha = 0.5,
                color = NA) + 
  geom_boxplot(notch = TRUE,
               width = 0.5) 
```

If the notches in the box plots do not overlap, the medians are probably significantly different.

# Dot Plot with PointRange Showing Means and 95% CI

```{r dotplot, message = FALSE, warning= FALSE}
p + geom_dotplot(binaxis = "y",
                 stackdir = "center", 
                 binwidth = .4, 
                 alpha = 0.5, 
                 color = NA) + 
  stat_summary(fun.data = "mean_cl_normal", 
               geom = "pointrange", 
               width = 0.1,
               color = c("royalblue","firebrick"),
               lwd = 1.5) 
```

# t-test write-up

```{r ttwriteup}
tt <- with(d, t.test(DV ~ IV, var.equal = TRUE))
tvalue <- tt$statistic %>% formatC(digits = 2, format = "f")
if (tt$p.value < 0.0005) {
    pvalue <- " < 0.001" 
  } else { 
    if (tt$p.value < 0.005)
      pvalue <- paste0(" = ",tt$p.value %>% formatC(digits = 3, format = "f"))
    else 
      pvalue <- paste0(" = ",tt$p.value %>% formatC(digits = 2, format = "f")) 
  }
df <- tt$parameter
```

This:

On average, Group 2 (m = `r rinline("dd$Mean[1]")`) scored significantly higher than Group 2 (m = `r rinline("dd$Mean[2]")`), `*t*`(`r rinline("df")`) = `r rinline("tvalue")`, `*p*` < `r rinline("pvalue")`.

Becomes:

On average, Group 2 (m = `r dd$Mean[1]`) scored significantly higher than Group 2 (m = `r dd$Mean[2]`), *t*(`r df`) = `r tvalue`, *p* `r pvalue`.

# Write-up with resusable functions

Automate your reports with frequently-used functions.

```{r ttreport}
t.report <- function(tt){
  tvalue <- tt$statistic %>% formatC(digits = 2, format = "f")
  pvalue <- tt$p.value %>% formatC(digits = 2, format = "f")
  if (round(tt$parameter, 0) == tt$parameter) {
    df <- tt$parameter
  } else {
    df <- formatC(digits = 2, format = "f")
  }
  if (tt$p.value < 0.0005) {
    pvalue <- " < 0.001" 
  } else { 
    if (tt$p.value < 0.005) {
      pvalue <- paste0(" = ",tt$p.value %>% formatC(digits = 3, format = "f"))
    } else {
      pvalue <- paste0(" = ",tt$p.value %>% formatC(digits = 2, format = "f"))
    }
    } 
  paste0("*t*(",df,") = ",tvalue, ", *p*", pvalue)
  }
```

This:

As seen in Table 1 and Figure 1, the mean of Group 2 is significantly higher than the mean of Group 2, `r rinline("t.report(tt)")`.

Becomes:

As seen in Table 1 and Figure 1, the mean of Group 2 is significantly higher than the mean of Group 2, `r t.report(tt)`.

```{r tablekable}
library(knitr)
#Make Table
kable(dd,
      digits = 2,
      caption = "Table 1: Descriptive Statistics for DV",
      align = "c")
```

```{r violinplotgroup, fig.cap="Figure 1: Violin Plot with Group Means and Standard Errors"}
ggplot(data = d, aes(y = DV, x = IV, fill = IV))  + 
  geom_violin(alpha = 0.5, color = NA) + 
  stat_summary(fun.data = "mean_cl_normal", 
               color = c("royalblue","firebrick"), 
               size = 1.1) 
```

The `papaja` package has many APA format functions that make life easier.

```{r papaja, eval=FALSE}
devtools::install_github("crsh/papaja")
```

[How not to worry about APA style](http://blog.efpsa.org/2015/06/09/how-not-to-worry-about-apa-style/)

# Write-up Files

[Markdown](http://my.ilstu.edu/~wjschne/444/IndependentSamplesWriteUp.Rmd)

[Word](http://my.ilstu.edu/~wjschne/444/IndependentSamplesWriteUp.docx)

[.pdf](http://my.ilstu.edu/~wjschne/444/IndependentSamplesWriteUp.pdf)

# *t*-test as Regression

$$Y_i=b_0+b_1 X_i + e_i$$

Where:

$Y_i$: The score on $\mathbf{Y}$ for person $i$

$b_0 = \mu_0$: The mean of the reference group

$b_1 = \mu_1 - \mu_0$: The difference of the two group means

$X_i$: Group membership for person $i$ (0 or 1)

$\mathbf{X}\sim B(1,p)$: $\mathbf{X}$ has a Bernoulli distribution with probability $p$.

$e_i$: The deviation of $Y_i$ from the group mean for person $i$

$\mathbf{e}\sim N(0,\sigma_e^2)$: $\mathbf{e}$ has a normal distribution with a mean of 0 and a variance of $\sigma_e^2$.

The distribution of $\mathbf{e}$ gives us both the normality assumption and the homogeneity of variance assumption.


```{r ttestregression, echo=F}
nIV <- (d$IV == "Group 2") * 1
cCol <- scales::alpha(c("royalblue","firebrick"), 0.5)
jIV <- nIV + rnorm(nrow(d),mean = 0,sd = 0.005) 
par(family = "serif", pty = "s")
plot(DV~jIV, 
     data = d, 
     col = cCol[nIV + 1],
     pch = 16, 
     xlab = "IV", 
     ylab = "DV", 
     las = 1, 
     xlim = c(-0.05,1.05),
     ylim = c(3,21),
     main = "Independent-Samples as Regression")
dd <- d %>% group_by(IV) %>% summarize(DV = mean(DV)) %>% ungroup() %>% select(DV)

lines(c(0,1),c(dd$DV))
text(0.5, dd$DV[1] + (dd$DV[2] - dd$DV[1])/2, bquote(italic(b[1]) == .(formatC(dd$DV[2] - dd$DV[1],2,format = "f"))),pos = 3,srt = 5)
```

Do *t*-test results accord with the regression results?

If $b_1=0$, we get a one-sample *t*-test:

$$\begin{align}Y_i&=b_0+0 X_i+ e_i\\&=\mu_0+e_i\end{align}$$



# Paired-Samples *t*-test {.SectionSlide}


# Import Paired Data

Variables | Label
:---------|:-------
`id`      | Case Number
`x_1`     | Time 1 score
`x_2`     | Time 2 score

```{r readd2}
d2 <- read.csv("http://my.ilstu.edu/~wjschne/444/pairedttest.csv")
kable(head(d2), 
      caption = "First 6 cases of d2", 
      digits = 2)
```

# Run Paired-Samples *t*-test

$$H_0: \mu_1=\mu_2$$

$$H_1: \mu_1\ne\mu_2$$

```{r pairedt}
with(d2, t.test(x_1, x_2, paired = TRUE))
```

# Plot Difference Scores&mdash;Histogram

```{r dplot}
d2d <- d2 %>% mutate(Difference = x_2 - x_1)

dplot <- ggplot(d2d, aes(x = Difference)) 
dplot + geom_histogram(binwidth = 1)
```

# Plot Difference Scores&mdash;Density Plot

```{r}
dplot + 
  geom_density(fill = "royalblue", 
               alpha = 0.5, 
               color = NA) + 
  geom_point(aes(y = 0),
             alpha = 0.5)
```

# Tidy Data

Many R packages assume that your data are structured in a way that [Hadley Wickham calls *tidy data*](http://vita.had.co.nz/papers/tidy-data.html).

The `tidyr` package has simple tools for data restructuring. For more complex restructring, we might need the `reshape2` package. Both packages are by Hadley Wickham (who also wrote `dplyr`, `ggplot2`, and many others).

## Definition of Tidy Data

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

[Data Management Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

[Great video on Data Wrangling](https://www.rstudio.com/resources/webinars/archives/)

[Slides](https://github.com/rstudio/webinars/raw/master/05-Data-Wrangling-with-R-and-RStudio/wrangling-webinar.pdf)

# Restructure data

```{r}
# Restructure data to make a tidy data set.
library(tidyr)
d2r <- d2 %>% 
  gather(key = "Time", 
         value = "x", 
         x_1:x_2) %>% 
  mutate(Time = extract_numeric(Time)) %>% 
  arrange(id, Time) %>% 
  mutate(Time = factor(Time))
# Print first 3 cases (6 rows)
kable(head(d2r), 
      digits = 2, 
      caption = "First three cases of d2r",
      align = "c")
```

# Line Plot with Errors

```{r}
ggplot(d2r, aes(Time, x)) + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) +
  stat_summary(fun.data = "mean_cl_normal",size = 1)
```


# Plot Paired-Points

```{r}
ggplot(d2r, aes(Time,x)) + 
  geom_violin(color = NA) +
  geom_point(aes(group = id, color = factor(id)), alpha = 0.3) +
  geom_line(aes(group = id, color = factor(id)), alpha = 0.3) + 
  stat_summary(fun.data = "mean_cl_normal",size = 5, geom = "point") +
  stat_summary(fun.y = "mean",size = 2, geom = "line", aes(group = 1)) +
  ylab("DV") 
```

# Paired-Samples as Regression

Assume that $b_1=1$

$\begin{align}Y_i&=b_0 + b_1 X_i + e_i\\&=b_0 + X_i + e_i\end{align}$

If $b_0\ne 0$, the means of $X$ and $Y$ are different.

```{r, echo = FALSE,fig.height = 7,fig.width = 7, dev = 'svg'}
par(family = "serif",pty = "s",xaxs = "i",yaxs = "i", mar = c(5,5,1,1))
plot(x_2 ~ x_1,data = d2, xlim = c(0,20), ylim = c(0,20), xlab = "X", ylab = "Y", pch = 16, col = rgb(0,0,1,0.5),las = 1, bty = "n")

mu1 <- mean(d2$x_1)
mu2 <- mean(d2$x_2)
b0 <- mu2 - mu1
abline(b0, 1)
abline(0, 1,lty = 3)
abline(v = mu1,lty = 3)
abline(h = mu2,lty = 3)
text(5,5 + b0,expression(italic(b[1] == 1)),srt = 45,pos = 1)
text(mu1,5,bquote(bar(italic(X)) == .(formatC(mu1,digits = 1,format = "f"))),pos = 4)
text(18,mu2,bquote(bar(italic(Y)) == .(formatC(mu2,digits = 1,format = "f"))),pos = 3)
Arrows(mu1,mu1,mu1,mu2,code = 3, arr.adj = 1, col = "firebrick",lwd = 1.5)
par(xpd = T)
text(0, b0, bquote(italic(b[0]) == .(formatC(b0,digits = 1,format = "f"))), col = "firebrick",  pos = 4)
Arrows(0,0,0,b0,code = 3, arr.adj = 1, col = "firebrick",lwd = 1.5)
par(xpd = F)

```


Inspired by [this paper](http://statland.org/AP/pairs.pdf)


