---
title: "MANOVA"
author: "W. Joel Schneider"
date: "Psy444: Multivariate Analysis"
params: 
  fast: TRUE
output: 
  slidy_presentation: 
    css: slidy.css
    fig_caption: yes
    highlight: kate
    widescreen: yes
    footer: <a href = 'http://my.ilstu.edu/~wjschne/444/Psy444FA2015.html'>Multivariate Analysis</a>
bibliography: Loaded.bib
csl: apa.csl
---

```{r setup,echo = FALSE,message = FALSE}
library(magrittr)
library(tables)
library(ggplot2)
library(knitr)
library(broom)
library(pander)
library(MVN)
library(scales)
library(car)
library(dplyr)

knit_hooks$set(familyserif = function(before, options, envir) {
    if (before) par(family = "serif")  
})
opts_chunk$set(dev = "svglite", familyserif = TRUE)
panderOptions("table.split.table",Inf)
panderOptions("round",2)
panderOptions("keep.trailing.zeros",TRUE)
panderOptions("table.emphasize.rownames", FALSE)
panderOptions("table.alignment.rownames", "left")
panderOptions("missing","")

pvalueAPA <- function(p, inline = FALSE, mindigits = 2, maxdigits = 3){
   fp <- function(x){
      p.round <- ifelse(x > 0.5 * 10 ^ (-1 * mindigits),mindigits,maxdigits)
  if (x > 0.5 * 10 ^ (-1 * p.round)) {
    paste0(ifelse(inline,"$p=", ""),
           sub(pattern = "0.", 
               replacement = ".", 
               formatC(x, p.round, format = "f")),
           ifelse(inline,"$", ""))
    } else {
      paste0(ifelse(inline, "$p<","<"),
             sub(pattern = "0.", 
                 replacement =  ".",
                 10 ^ (-1 * maxdigits)),
             ifelse(inline,"$",""))
    }
  }
  sapply(p, fp)
}

rinline  <-  function(code) {
  sprintf('``` `r %s` ```', code)
}

themeMod <- theme_grey() + 
  theme(text = element_text(family = "serif", size = 10),
        legend.position = "none",
        axis.title.x = element_text(vjust = -0.75),
        axis.title.y = element_text(vjust = 1.75))
theme_set(themeMod)
# Custom colors
myFill <- function(...){
  scale_fill_manual(values = alpha(c("royalblue", "firebrick", "darkorchid"),0.5),...)
}
scale_fill_discrete <- myFill
myColor <- function(...){
  scale_colour_manual(values = c("royalblue", "firebrick", "darkorchid"),...)
}
scale_colour_discrete <- myColor


manovaTable <- function(m, caption = "MANOVA Table", test = "Pillai"){
  require(broom)
  require(pander)
  require(dplyr)
  testName <- switch(test,
                     Pillai = "Pillai's Trace", 
                     Wilks = "Wilks' $\\Lambda$",
                     `Hotelling-Lawley` = "Hotelling-Lawley's Trace",
                     Roy = "Roy's Largest Root")
  m %>% manova %>% 
    tidy(test = test) %>% 
    mutate(p.value = pvalueAPA(p.value),
           term = gsub(x = term, pattern = ":"," $\\\\times$ ") ) %>% 
    set_colnames(c("Factor",
                   "df",
                   testName,
                   "$F$",
                   "$df_1$",
                   "$df_2$",
                   "$p$")) %>% 
    pander(caption = caption, 
           round = 2, 
           split.tables = Inf, 
           style = "rmarkdown", 
           emphasize.rownames = FALSE, 
           keep.trailing.zeros = TRUE, 
           table.alignment.rownames = "left", 
           missing = "")
}
manovaUnivariateTable <- function(m, caption = "Univariate Analyses"){
  require(broom)
  require(pander)
  require(dplyr)
  require(car)
  m %>% tidy %>% 
    mutate(response = ifelse(response == Lag(response),
                             NA,
                             response),
           p.value = pvalueAPA(p.value),
           term = gsub(x = term, pattern = ":"," $\\\\times$ "))  %>% 
    set_colnames(c("Outcome",
                   "Predictor",
                   "$b$",
                   "$\\sigma_b$",
                   "$t$",
                   "$p$")) %>% 
    pander(caption = "Univariate Analyses")
}


```

# ANOVA on steroids

|   |1 Response Variable|2 or More  Response Variables|
|---|:----------------:|:---------------------------:|
|*1 Dichotomous Predictor* |*t*-test|Hotelling's *T^2^*|
|*1 or More Polytomous Predictors*|ANOVA|MANOVA|
|*1 or More Continuous Predictors*|Multiple Regression|Canonical Correlation Analysis|

```{r MakeData, include=FALSE}
set.seed(1)
n <- 300
main <- sample(c(0,1), n * 2, TRUE) %>% matrix(ncol = 2)
x <- cbind(rep(1,n), main, main[,1] * main[,2]) 
b <- cbind(c(10,1,-1,0),
           c(10,1,0,-3),
           c(10,1,-1,0))
yhat <- x %*% b
eSD <- c(3,3, 3)
ecor <- matrix(c(1.0,0.4,0.6,
                 0.4,1.0,0.5,
                 0.6,0.5,1.0), 3)
ecov <- lavaan::cor2cov(ecor, eSD)
y <- yhat + mvtnorm::rmvnorm(n, sigma = ecov)
d <- data.frame(x,y) %>% set_colnames(c("Intercept","A","B","AB","Y1","Y2","Y3")) %>% select(-Intercept, -AB) %>% mutate(A = factor(A, labels = c("A1","A2")), B = factor(B, labels = c("B1","B2")))
write.csv(d,"manova.csv", row.names = FALSE)
```

# MANOVA in R

# Descriptives

```{r}
head(d) %>% pander

tabular((Factor(A, levelnames = c("$A_1$","$A_2$")) + 1) * 
        (Factor(B, levelnames = c("$B_1$","$B_2$")) + 1) ~ 
         Heading("N") * 1 + 
          (Y1 + Y2 + Y3) * 
        (Heading("Mean") * mean + 
         Heading("SD") * sd), 
        data = d) %>% pander(emphasize.rownames = F)
```




# Plotting MANOVA results

```{r}
ggplot(d, aes(A,Y1)) + 
  geom_violin(col = NA) + 
  stat_summary(fun.data = "mean_cl_normal") + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) +
  facet_grid(. ~ B)

ggplot(d, aes(A,Y2)) + 
  geom_violin(col = NA) + 
  stat_summary(fun.data = "mean_cl_normal") + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) +
  facet_grid(. ~ B)

ggplot(d, aes(A,Y2)) + 
  geom_violin(col = NA) + 
  stat_summary(fun.data = "mean_cl_normal") + 
  stat_summary(fun.y = "mean", geom = "line", aes(group = 1)) +
  facet_grid(. ~ B)
```


# Multivariate Effects

```{r}
m <- lm(cbind(Y1,Y2,Y3) ~ A * B, d)

m %>% manova %>% summary
m %>% manova %>% summary(test = "Wilks")
m %>% manova %>% summary(test = "Hotelling-Lawley")


```

Roy's Largest Root

Measures distances on the first canonical root.
```{r}
m %>% manova %>% summary(test = "Roy")

```

# Multivariate Table

See `manovaTable` function in the setup chunk

```{r}
m %>% manova %>% tidy 

manovaTable(m)
```


# Univariate Effects

```{r}
m %>% summary

m %>% summary.aov
```

# Univariate Table

See `manovaUnivariateTable` function in the setup chunk

```{r unitable}
m %>% tidy

manovaUnivariateTable(m)

```

# Convenient Display

```{r manovadisplay}
m %>% car::Manova() %>% summary
```


# Multivariate Outler

```{r multioutliers}
e <- mvOutlier(m$residuals)
```


# Assumptions

ANOVA | MANOVA
:----------|:--------
Normality | Multivariate Normality
Independence | Independence (Dependent measures may be correlated)
Homogeneity of Variance $(\sigma_1^2=\sigma_2^2)$|Homogeneity of Covariance $\boldsymbol{\Sigma_1=\Sigma_2}$

# Univariate Assumptions

```{r uniassumptions}
lm(Y1 ~ A * B, d) %>% plot(which = 1)
lm(Y1 ~ A * B, d) %>% plot(which = 2)
lm(Y1 ~ A * B, d) %>% plot(which = 3)
lm(Y1 ~ A * B, d) %>% plot(which = 4)
lm(Y1 ~ A * B, d) %>% plot(which = 5)
lm(Y1 ~ A * B, d) %>% plot(which = 6)
lm(Y2 ~ A * B, d) %>% plot(which = 1)
lm(Y2 ~ A * B, d) %>% plot(which = 2)
lm(Y2 ~ A * B, d) %>% plot(which = 3)
lm(Y2 ~ A * B, d) %>% plot(which = 4)
lm(Y2 ~ A * B, d) %>% plot(which = 5)
lm(Y2 ~ A * B, d) %>% plot(which = 6)

d$Y3[1] <- 500

lm(Y3 ~ A * B, d) %>% plot(which = 1)
lm(Y3 ~ A * B, d) %>% plot(which = 2)
lm(Y3 ~ A * B, d) %>% plot(which = 3)
lm(Y3 ~ A * B, d) %>% plot(which = 4)
lm(Y3 ~ A * B, d) %>% plot(which = 5)
lm(Y3 ~ A * B, d) %>% plot(which = 6)

d$Y3[1] <- NA
lm(Y3 ~ A * B, d) %>% plot(which = 1)
lm(Y3 ~ A * B, d) %>% plot(which = 2)
lm(Y3 ~ A * B, d) %>% plot(which = 3)
lm(Y3 ~ A * B, d) %>% plot(which = 4)
lm(Y3 ~ A * B, d) %>% plot(which = 5)
lm(Y3 ~ A * B, d) %>% plot(which = 6)

```


# Multivariate Normality of Residuals

```{r multivariatenormality}
library(MVN)
mardiaTest(m$residuals, qqplot = TRUE)
mvnPlot(mardiaTest(m$residuals[,c(1,2)]))
mvnPlot(mardiaTest(m$residuals[,c(2,3)]))
mvnPlot(mardiaTest(m$residuals[,c(1,3)]))
```

# Homogeneity of Covariance

Box's M tests the equality of covariance matrices. It is extremely sensitive to normality violations.

```{r homogeneitycovariance}
biotools::boxM(d %>% select(Y1, Y2, Y3), d$A)
biotools::boxM(d %>% select(Y1, Y2, Y3), d$B)
```

# References

In making these slides, I relied heavily on:

Schumacker, R. E. (2016). *Using R with multivariate statistics*. Thousand Oaks, CA: Sage.
